A New Role for RNA G-quadruplexes in Aging and Alzheimer’s Disease

Version 1

Abstract

Abstract

Introduction

As the world population ages, new molecular targets in aging and Alzheimer’s Disease (AD) are needed to combat the expected influx of new AD cases. Until now, the role of RNA structure in aging and neurodegeneration has largely remained unexplored. METHODS: In this study, we examined human hippocampal postmortem tissue for the formation of RNA G-quadruplexes (rG4s) in aging and AD.

Results

We found that rG4 immunostaining strongly increased in the hippocampus with both age and with AD severity. We further found that neurons with accumulation of phospho-tau immunostaining contained rG4s, that rG4 structure can drive tau aggregation, and that rG4 staining density depended on APOE genotype in the human tissue examined.

Discussion

Combined with previous studies showing the dependence of rG4 structure on stress and the extreme power of rG4s at oligomerizing proteins, we propose a model of neurodegeneration in which chronic rG4 formation drives proteostasis collapse. These morphological findings suggest that further investigation of RNA structure in neurodegeneration is a critical avenue for future treatments and diagnoses.

Introduction

As the aged population increases, Alzheimer’s Disease (AD) and AD-related dementias (ADRDs) will become one of the world’s largest medical and economic crises due to the lack of early diagnostic methods and durable therapies1, 2. Age is the highest risk factor for AD, as disease prevalence increases exponentially as a function of age, and close to 40% of older adults over the age of 80 experience AD or ADRDs3, 4. The most important genetic risk factor for developing late onset AD (LOAD) is the Apolipoprotein E (APOE) genotype, at least in Caucasians, where carriers of one or two copies of the APOE4 allele have a substantial increase in the risk for developing LOAD when compared to carriers of the APOE2 or 3 alleles4, 5. Beyond APOE, recent meta-analyses combining genome-wide association studies (GWAS) have expanded the number of AD-risk loci6, but most of the disease-associated variants reside in non-protein coding regions of the genome, making it difficult to elucidate how they affect AD susceptibility and pathogenesis.

The exact mechanisms behind these two predominant AD risk factors remains a mystery but both are widely associated with increased abnormal protein aggregation, which is the most abundantly observed pathological hallmark of AD pathogenesis. AD and ADRDs all have an abnormal protein aggregation profile that includes: 1) extracellular amyloid senile plaque deposition occurring at least 5-10 years before mild cognitive impairments (MCI) are observed and, 2) aggregates of phosphorylated Tau (p-Tau) forming intracellular neurofibrillary tangles (NFTs) near MCI onset7, 8. Novel biomarker studies have shown that hyperphosphorylated Tau species exist in plasma prior to onset of dementia symptoms, with specific phosphorylation sites seeming to affect the affinity of binding of the Tau protein to microtubules 9. With advancing age, protein aggregation is accelerated by the accumulation of reactive oxygen species (ROS) as well as the onset of neuroinflammation in the brain10–12. Having APOE4 was shown to significantly increase accumulation of amyloid and NFTs in Caucasians, and also increases inflammation and reactive oxygen species in the human brain5, 13. The effects of these AD risk factors have been studied mostly at the protein level in the brain, but not yet at RNA level, where structure and function have a myriad of important roles in numerous biological processes.

RNA higher order structures of particular interest are RNA guanine-rich G-quadruplexes (rG4s)14. G-quadruplexes are secondary structures that can fold under physiological conditions by guanine-rich DNA and RNA molecules and occur in the genomes and transcriptomes of many species, including humans15. rG4s are four-stranded helices formed by guanine tetrads through Hoogsteen base-pairing, with a monovalent cation channel running down the middle of the tetrads (Fig. 1a)16. G4s are potentially involved in the regulation of DNA and RNA processes, such as replication, transcription, translation, RNA localization, and degradation17–19. Recently, rG4s were shown to form as a function of stress in eukaryotic cells20. Under non-stress conditions, rG4s were largely unfolded, but upon introduction of stress (including ROS stress) the rG4s preferentially folded and remained folded until the cessation of stress21. This observation raises the possibility that rG4s could preferentially fold under the increased and chronic stress conditions of aging and AD in the human brain, leading to abnormally high levels of rG4 formation in older adults or in adults with protein misfolding diseases22. RNA has been known to be included in AD-related aggregates of both amyloid and NFTs in patients with AD for decades23, and a recent sequencing study of nucleic acids embedded in AD aggregates showed an enrichment in potential rG4-forming sequences24.

a, (left) Representative G-quadruplex tetrad structure with Hoogsteen base pairing stabilized by a cation at the center. (right) G-quadruplex structure with stacked tetrads. b, Demographics table for individuals included in this study. c, Immunofluorescence showing BG4 staining (red) in brain sections of human hippocampus from four control individuals of increasing age ranging 30-91 years old (Case 15, 13, and 8, respectively) all with Braak stage 0. d, Correlation of percent area covered by BG4 fluorescence in the outer-molecular layer versus age of individual.

It has recently been shown that rG4s can greatly enhance protein oligomerization25, 26. Moreover, rG4s can bind to and greatly enhance the phase transition of tau27. Similarly, Zwierzchowski-Zarate et al explored the effects of different RNA sequences on tau aggregation in a cellular model28. In their study, several sequences caused an increase in tau aggregation, but the highest level of tau aggregation was achieved with a GGGC repeat sequence28. However, the structure of the RNAs, including possible rG4 formation, was not explored or speculated upon. We therefore performed CD spectroscopy to determine if the sequence associated with the highest levels of tau fibrillation might be forming rG4 secondary structures. The CD spectra show that this RNA forms parallel rG4 structures in vitro (Supplementary Data Fig. 1).

Based on these previous in vitro and cellular studies, an increased presence of rG4s in the brain could contribute to AD-associated aggregation. However, the presence of rG4s in the human brain or other human tissues has not been shown previously, nor how it relates to aging or AD pathology. This was the focus of the current study.

Methods

Human brain tissue

This work was approved by the University of Colorado Institutional Review Board (IRB) and the study was performed in accordance with the ethical standards of the 1964 Declaration of Helsinki and its later amendments. However, the use of postmortem brain tissue for research is exempt according to federal IRB regulations and does not require IRB review and approval. All efforts were made to consider diversity, equity and inclusion (DEI) in the selection of brain tissues for this study. The tissues included in this study were obtained from the Medical University of South Carolina (MUSC) brain bank (Dr. Eric Hamlett). Postmortem consent was obtained from the next of kin, and the brain was rapidly removed. Fig. 1b shows the demographics for the 21 cases used for this study. The tissue was sliced in 1 cm coronal slices and subjected to free-floating fixation in 4% paraformaldehyde in phosphate buffered saline (PBS) for 72 hours. The fixed tissue was then transferred to a 30% sucrose in phosphate buffer (PB), followed by transfer to a cryoprotection solution in PBS (30% Glycerol, 30% Ethylene Glycol and 40% PO4 Buffer) and stored in −20° C until dissection. Anatomical blocks from 19 brain regions were dissected from the 1 cm slabs, embedded in paraffin and cut at a thickness of 5µm on a Microm Microtome (Thermo Fisher Scientific Inc., Waltham, MA) and stained for routine neuropathological staging according to the NIA-AA Revised Criteria for Diagnosis and Staging of Alzheimer’s Disease 29.

Immunofluorescence

Paraffin-embedded sections were deparaffinized in xylene (RPI, CAT#111056) and rehydrated using wash steps at: 100%, 95%, 70%, 50% ethanol followed by distilled water. Heat-induced antigen retrieval was performed using 0.05% citraconic acid at pH 7.2 for 20 minutes in a steamer. Slides were then washed with a saline solution buffered with tris(hydroxymethyl) aminomethane (Tris, TBS) and blocked with 0.25% triton-X and 5% normal serum for 1 hour. After blocking, slides were rinsed again with TBS and incubated with 1X True Black Lipofuscin Autofluorescence Quencher (Biotium CAT#23007) diluted in 70% ethanol. Slides were washed 3 times with TBS then incubated with primary antibodies listed in Supplementary Data Table 1 overnight, followed by Goat anti-mouse AlexaFluor 555 (Thermo Scientific #A32727) or Donkey anti-rabbit Alexafluor 488 (Invitrogen #A-21206) at 1:500 dilution for 1 hour. Slides were then washed 3 times with TBS and cover-slipped using mounting media containing DAPI (Thermo Scientific, CAT#P36931). Staining controls included omission of primary or secondary antibodies (Supplementary Data Fig. 2). In Fig. 3b, we performed sequential double labeling, first with BG4, and then with rRNA antibody at 1:250 (Abcam #ab17119).

In addition, we examined whether Braak stage correlated with aging itself (Supplementary Data Fig. 3). Although not significant at the 0.05 level, there was a weak correlation between Braak and Age. Braak staging is a paradigm used to describe density and regional spread of neurofibrillary tangles (NFTs) in the Alzheimer brain and includes levels 0-VI7. CERAD (Consortium to Establish a Registry for Alzheimer’s Disease) is a staging paradigm for Alzheimer’s disease that takes into account amyloid plaque formation30.

Imaging and Statistical methods

Slides were imaged on a confocal microscope at 10X, 20X and 60X magnification at wavelengths 355, 488, and 555. Images used for quantification were taken at constant microscope and intensity settings. All image quantification was performed using Fiji ImageJ. Quantification was performed by a researcher who was blinded to the neuropathological diagnosis of individual cases and with no knowledge of Braak Stage or Age for the case being quantified. Supplemental Table 2 presents side by side comparisons of quantification from 2 researchers who were both blinded to group diagnosis and individual cases. Data used in the paper are highlighted. One data point with too much variation between quantifications was removed to increase rigor. Background was subtracted using a rolling ball radius. Images were then divided into 3 regions of interests (ROI); OML, DG and CA4. (Supplementary Data Fig. 4) Images were submitted to an auto-threshold and particle’s analyzed to get a percent area measurement of BG4 staining. These were graphed and linearly fit using Kaleidagraph using Pearson fit. Spearman and Pearson R and p-values are reported for 18 cases for age correlation and 18 cases for Braak correlation. An unpaired T-test with unequal variance was performed for the two groups E2/E3 and E3/E3 vs E3/E4 and E4/E4. When the singular African American case was removed from the APOE quantification, the t-probability remained highly significant at 0.0045 vs 0.002. All quantification was performed on 10X images obtained from an Olympus microscope.

Circular Dichroism

CD spectra were obtained using a Jasco J-1100 circular dichroism at 25°C. Sequence was resuspended in 10mM potassium phosphate pH 7.5 buffer and diluted to 12.5 µM (per strand) RNA. The CD measurement was taken from 300 nm to 210 nm at 1nm intervals using a 50 nm/min scanning speed. The sequence used was 5’- CGGGCGGCGGGGGGGCCCGGGCGGCGGGGGGGCCCGGGCG-3’ 28. Parallel structure was determined using previously described quantification methods for quadruplex topology31.

Results

Formation of rG4s in aging

To determine if rG4s form in the human brain, we used the well-characterized single-chain BG4 anti-quadruplex antibody (Sigma-Aldrich MABE917, antibody null controls in Supplementary Data Fig. 2) to stain human hippocampal postmortem tissue of different ages under conditions that strongly favor cytoplasmic rG4 identification using triton-X fixation in cells 32, 33. The tissues were obtained in collaboration with the brain bank at the Medical University of South Carolina (MUSC). The study included 21 cases, male and female, of different ages and with or without a clinical or neuropathological diagnosis of AD (see Fig. 1b). All cases had received a neuropathological examination including assessment of Braak and CERAD stages according to the updated NIA criteria except in those cases in which no pathology was noted 7, 34.

First, the rG4 staining profiles were compared between different ages, with an age range of 30-92 years of age (Fig. 1c). We discovered a significant age-related increase in rG4 immunostaining, with the densest staining observed in older individuals (Fig. 1c,d). The largest difference in staining between young and older cases was observed in the outer molecular layer (OML) and in the hilar region, CA4 (see Fig. 1c, legend arrows). The strongest overall staining for BG4 was found in the dentate gyrus granular cell layer (DG), but with less observable age-related increases. Staining density measurements confirmed these findings and demonstrated a highly significant increase in BG4 staining in the oldest cases studied within the OML (Fig. 1d) and the CA4 regions of cornu ammonis (Supplementary data Fig. 5a), confirming increased intracellular BG4 accumulation as a function of age. Plotting BG4 percent area versus age confirmed the observed findings and showed correlation (Spearman R= 0.49) with high significance (p= 0.038 in the OML region (Fig. 1d). Positive and significant Pearson but not Spearman correlations were observed between age and BG4 area in the CA4 region, Supplementary data Fig. 5a), and non-significant correlations were observed in the dentate granule cell layer. No significant correlation was found with BG4 stain and postmortem interval (PMI) (Spearman R = −0.18, p = 0.49 see Supplementary Data Fig. 6). In total, these data show that rG4s increased as a function of age in the human brain. We observed multiple outliers in both older and younger individuals due to Braak stage. For example, an 85-year-old individual with Braak stage 0, or 59-year-old individual with Braak stage 5.5 were outliers as a function of age. This led us to analyze whether rG4 formation increased with Braak stage, a measure of AD severity.

Formation of rG4s in AD

Next, we examined whether the formation of rG4s was associated with AD neuropathology. Braak staging is a neuropathological staging method that takes into account the density of neurofibrillary tangles (NFTs) both within each brain region and also from region to region30, with higher numerical values representing greater AD severity7. Like in the case of aging, increased BG4 staining was readily apparent in AD cases, especially in those cases associated with a higher Braak stage (Fig. 2a). In the OML, plotting BG4 percent area versus Braak stage demonstrated a strong correlation (Spearman R= 0.72) with highly significantly increased BG4 staining with higher Braak stages (p = 0.00086) (Fig. 2b). Positive significant correlations (R = 0.52, p = 0.028) between Braak stage severity and BG4 staining were also observed within the CA4 region (Supplementary data fig. 4). Interestingly, the OML region is highly involved in AD pathology, and is a hippocampal region where AD pathology is often observed,35 suggesting that the presence of rG4s in this particular brain region might have significance for amyloid as well as NFT formation during the AD disease process. In addition, a prominent loss of synapses in AD is found in the OML, suggesting a potential close connection with layer 2 of the entorhinal cortex36.

a, Immunofluorescence showing BG4 staining (red) in brain sections of human hippocampus with increasing Braak stage (1-6) associated with AD severity (Case 16, 11, 10, and 8, respectively). All scale bars are 300 µm. b, Correlation of percent area BG4 coverage in the outer molecular layer versus severity of Braak stage. c, Quantification of percent area BG4 coverage by population with E2/E3 and E3/E3 versus E3/E4 and E4/E4 APOE alleles.

If rG4 formation is more prevalent in AD than in non-AD cases and directly involved in AD pathology, we would also predict that rG4 formation would be more prevalent in APOE genotypes that are predisposed for higher LOAD risk, i.e. patients with one or two copies of the APOE4 genotype. Indeed, there was a significant (t-probability 0.002) increase in BG4 staining levels in the OML in postmortem cases with one or more APOE4 alleles compared to those with only APOE2 or APOE3 alleles (Fig. 2c). This result, although correlative, is consistent with a previous study performed in vitro in which the RNA sequences found in aggregates in different APOE genotype cell lines were different24.

Patterns of rG4 localization

Images obtained at high magnification using a confocal microscope (Olympus) showed that a large amount of the BG4 immunostaining was localized in the cytoplasm near the nucleus in a punctate pattern, reminiscent of known rG4 formation in vitro and in cell lines in rRNA ribosomal extensions33. We therefore also stained with an anti-rRNA antibody to determine whether this was the case in human tissue as well. As expected, the stains overlapped (Fig. 3b, Fig. S7), showing that at least a large percentage of the observed BG4 signal arose from rG4s and not nuclear DNA G4s. This result also confirms that previously observed rG4 formation in ribosomes33 also occurs in human brain tissue.

Cellular types and localization of rG4s, all scale bars are 50 µm. a, Immunofluorescence showing BG4 staining patterns (red) in human hippocampal tissue of case 17, a 53-year-old individual with Braak stage 1.5, and case 4, a 79-year-old individual with Braak stage 5 AD pathology (right). Cell nuclei stained with DAPI (blue). b, Co-stain of rRNA (left, yellow green) and BG4 (right, red), in the hippocampal CA4 region, an 80-year old individual with early stage tauopathy, demonstrating significant but not complete overlap between BG4 and rRNA. c, BG4 staining (red) with cell type markers (green) of oligodendrocytes (SOX10, case 8), neurons (NEUN, case 4), microglia (IBA1, case 18) and astrocytes (GFAP, case 18). BG4 is most prevalent in oligodendrocytes, neurons, and astrocytes. BG4 does not strongly colocalize with microglia (IBA1).

To further explore the formation of rG4s in the human hippocampus, we examined the staining pattern in different cell types based on specific antibody markers. Co-staining with markers for the most common brain cell types, rG4s were clearly abundant in neurons, oligodendrocytes, and astrocytes (Fig. 3c). rG4s exhibited no apparent colocalization with microglia (Fig. 3c). This pattern is especially noticeable in the double labeling with SOX10 (Fig. 3c), which clearly shows the nuclear oligodendrocyte marker SOX1037 staining in proximity to the cytoplasmic BG4 immunostaining.

We co-stained BG4 with different p-tau epitope antibodies to determine whether pathological aggregation of tau proteins co-occurred with BG4 immunostaining (Fig. 4). Indeed, we found that BG4 immunostaining co-localized with hyper-phosphorylated tau immunostaining in neurons located in the CA1-4 regions of the hippocampus (Fig. 4). We used two different antibodies directed against the serine 396 site (S396) and the threonine 231 (T231) to investigate co-localization with different p-Tau epitopes. Quantifying the percentage of cells co-stained with BG4 and pTau revealed that 85% of the S396 p-tau positive cells also had positive BG4 staining, and 82% of the T231 p-tau positive cells also had positive BG4 staining.

Colocalization of rG4s with p-tau and neurodegeneration model.

a, Immunofluorescence of human hippocampus from 2 older AD individuals (case 8 and case 21)showing BG4 (red), DAPI (blue), and p-tau (green). b, Proposed model relating chronic rG4 formation and protein aggregation in aging and neurodegeneration.

Discussion

In this study, we demonstrated that rG4s form preferentially as a function of age and AD pathology in neurons and glial cells the human postmortem brain. While the granule cell layer in the dentate gyrus exhibited the most BG4 immunostaining overall, this was not an area most changed with aging and AD, as considerably greater changes were observed in the OML and the CA4 region of the hippocampus. rG4s were primarily found in the cytoplasm near the nucleus of neurons, astrocytes, and oligodendrocytes and were also found in neurons that contained regions of positive staining against two different epitopes of p-tau. The BG4 staining levels were also impacted by APOE status, with the presence of one or two APOE4 alleles associated with higher BG4 levels in the OML region. The OML is an interesting region of the hippocampus from the pathology standpoint, since this is the region of the hippocampus exhibiting the most prominent synaptic loss early in the disease process36, 38. We have recently demonstrated a significant loss of presynaptic components of the OML in AD38, strengthening the findings observed herein and suggesting a particular vulnerability of this region of the hippocampus to AD progressive pathology.

Combining these observations with those previously made on the reliance of rG4 structure on cellular stress and the roles of rG4s in protein oligomerization and tau aggregation in vitro, we can propose a model of how rG4s form and could perpetuate the effects of AD (Fig. 4c). As ROS and other protein stresses accumulate with age and according to APOE genotype, rG4s form. Although rG4s can rescue protein folding and could release proteins when unfolded at stress cessation39, under persistent stress proteins RNA would not be released from the G-quadruplexes, leading to increased protein oligomerization and aggregation26. This function has previously been suggested for RNA in biomolecular condensates more generally40. The acceleration of protein oligomerization and aggregation by the rG4s could then contribute to a vicious cycle of increased aggregation, leading to synaptic loss as described recently for the OML38. Recent in vitro work confirms an interaction between rG4 formation and p-Tau aggregation specifically41. In vitro experiments focused on the specific mechanisms for the rG4 influence on Tau phosphorylation and/or aggregation will be a future focus following these preliminary observations. Of note, the previous studies showing that the fixation conditions used here strongly favor rG4 identification instead of DNA G4 identification were performed in cells, and so it is possible that in tissue some DNA staining could also occur, however this is not suggested by the co-stain with rRNA (Fig. 3b).

Of note, the accumulation or rG4s in neurodegeneration could have roles in addition to those of protein aggregation. RNA transport dysregulation has also been a common factor identified in neurodegeneration42, and it could be possible that the rG4 build-up observed here could be in part a product of this process or could lead to it. It has also been proposed that the RNA transport dysregulation and protein aggregation processes could be linked 43. This remains an area of needed future investigation.

This work raises the possibility of using rG4s as potential drug targets in AD, or given the high correlation with AD severity, as early AD biomarkers. Future work can focus on ameliorating AD symptoms through rG4 binding molecules or detecting brain-derived rG4s as disease biomarkers for diagnosis. Many different rG4-binding molecules have already been created and shown to bind in human cells44, and provide a set of promising lead compounds for potential AD treatment.

Although here we have examined only AD, there are many other aging-related diseases featuring protein aggregation, including Parkinson’s Disease, ALS, Fragile X syndrome, and Huntington’s Disease 45. Notably, all of these diseases now have significant evidence showing rG4s as interactors with the aggregating proteins 46–58, and in the case of Fragile X, a recent study demonstrated that rG4s could contribute to its aggregation and neurotoxicity59. In addition, a recent manuscript by Raguseo et al., showed an increased prevalence of G4-structures in C9orf72 mutant human motor neurons in ALS/FTD when compared to healthy motor neurons, and direct involvement of the rG4 complex in the aggregation process60. These recent findings illustrate the need to explore rG4 complexes and their role in neurodegeneration further. The results here combined with these previous studies suggest that the formation of rG4s could be a common mechanism of neurodegeneration.

eLife Assessment

The current human tissue-based study provides convincing evidence correlating hippocampal expressions of RNA guanine-rich G-quadruplexes with aging and with Alzheimer's Disease presence and severity. The results are important and hold promise for deeper understanding of AD's pathogenesis and potential new therapeutic strategies.

[Editors' note: this paper was reviewed by Review Commons.]

Reviewer #1 (Public review):

This is an interesting manuscript where the authors systematically measure rG4 levels in brain samples at different ages of patients affected by AD. To the best of my knowledge this is the first time that BG4 staining is used in this context and the authors provide compelling evidence to show an association with BG4 staining and age or AD progression, which interestingly indicates that such RNA structure might play a role in regulating protein homeostasis as previously speculated. The methods used and the results reported seems robust and reproducible. There were two main things that needed addressing:

(1) Usually in BG4 staining experiments to ensure that the signal detected is genuinely due to rG4 an RNase treatment experiment is performed. This does not have to be extended to all the samples presented but having a couple of controls where the authors observe loss of staining upon RNase treatment will be key to ensure with confidence that rG4s are detected under the experimental conditions. This is particularly relevant for this brain tissue samples where BG4 staining has never been performed before.

(2) The authors have an association between rG4-formation and age/disease progression. They also observe distribution dependency of this, which is great. However, this is still an association which does not allow the model to be supported. This is not something that can be fixed with an easy experiment and it is what it is, but my point is that the narrative of the manuscript should be more fair and reflect the fact that, although interesting, what the authors are observing is a simple correlation. They should still go ahead and propose a model for it, but they should be more balanced in the conclusion and do not imply that this evidence is sufficient to demonstrate the proposed model. It is absolutely fine to refer to the literature and comment on the fact that similar observations have been reported and this is in line with those, but still this is not an ultimate demonstration.

Comments on current version:

The authors have now addressed my concerns.

Reviewer #2 (Public review):

RNA guanine-rich G-quadruplexes (rG4s) are non-canonical higher order nucleic acid structures that can form under physiological conditions. Interestingly, cellular stress is positively correlated with rG4 induction.

In this study, the authors examined human hippocampal postmortem tissue for the formation ofrG4s in aging and Alzheimer Disease (AD). rG4 immunostaining strongly increased in the hippocampus with both age and with AD severity. 21 cases were used in this study (age range 30-92).

This immunostaining co-localized with hyper-phosphorylated tau immunostaining in neurons. The BG4 staining levels were also impacted by APOE status. rG4 structure was previously found to drive tau aggregation. Based on these observations, the authors propose a model of neurodegeneration in which chronic rG4 formation drives proteostasis collapse.

This model is interesting, and would explain different observations (e.g., RNA is present in AD aggregates and rG4s can enhance protein oligomerization and tau aggregation).

Main issue from the previous round of review:

There is indeed a positive correlation between Braak stage severity and BG4 staining, but this correlation is relatively weak and borderline significant ((R = 0.52, p value = 0.028). This is probably the main limitation of this study, which should be clearly acknowledged (together with a reminder that "correlation is not causality"). Related to this, here is no clear justification to exclude the four individuals in Fig 1d (without them R increases to 0.78). Please remove this statement. On the other hand, the difference based on APOE status is more striking.

Comments on current version:

The authors have made laudable efforts to address the criticisms I made in my evaluation of the original manuscript.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

This is an interesting manuscript where the authors systematically measure rG4 levels in brain samples at different ages of patients affected by AD. To the best of my knowledge this is the first time that BG4 staining is used in this context and the authors provide compelling evidence to show an association with BG4 staining and age or AD progression, which interestingly indicates that such RNA structure might play a role in regulating protein homeostasis as previously speculated. The methods used and the results reported seems robust and reproducible. There were two main things that needed addressing:

(1) Usually in BG4 staining experiments to ensure that the signal detected is genuinely due to rG4 an RNase treatment experiment is performed. This does not have to be extended to all the samples presented but having a couple of controls where the authors observe loss of staining upon RNase treatment will be key to ensure with confidence that rG4s are detected under the experimental conditions. This is particularly relevant for this brain tissue samples where BG4 staining has never been performed before.

(2) The authors have an association between rG4-formation and age/disease progression. They also observe distribution dependency of this, which is great. However, this is still an association which does not allow the model to be supported. This is not something that can be fixed with an easy experiment and it is what it is, but my point is that the narrative of the manuscript should be more fair and reflect the fact that, although interesting, what the authors are observing is a simple correlation. They should still go ahead and propose a model for it, but they should be more balanced in the conclusion and do not imply that this evidence is sufficient to demonstrate the proposed model. It is absolutely fine to refer to the literature and comment on the fact that similar observations have been reported and this is in line with those, but still this is not an ultimate demonstration.

Comments on current version:

The authors have now addressed my concerns.

We thank the reviewer for their support!

Reviewer #2 (Public review):

RNA guanine-rich G-quadruplexes (rG4s) are non-canonical higher order nucleic acid structures that can form under physiological conditions. Interestingly, cellular stress is positively correlated with rG4 induction.

In this study, the authors examined human hippocampal postmortem tissue for the formation ofrG4s in aging and Alzheimer Disease (AD). rG4 immunostaining strongly increased in the hippocampus with both age and with AD severity. 21 cases were used in this study (age range 30-92).

This immunostaining co-localized with hyper-phosphorylated tau immunostaining in neurons. The BG4 staining levels were also impacted by APOE status. rG4 structure was previously found to drive tau aggregation. Based on these observations, the authors propose a model of neurodegeneration in which chronic rG4 formation drives proteostasis collapse.

This model is interesting, and would explain different observations (e.g., RNA is present in AD aggregates and rG4s can enhance protein oligomerization and tau aggregation).

Main issue from the previous round of review:

There is indeed a positive correlation between Braak stage severity and BG4 staining, but this correlation is relatively weak and borderline significant ((R = 0.52, p value = 0.028). This is probably the main limitation of this study, which should be clearly acknowledged (together with a reminder that "correlation is not causality"). Related to this, here is no clear justification to exclude the four individuals in Fig 1d (without them R increases to 0.78). Please remove this statement. On the other hand, the difference based on APOE status is more striking.

Comments on current version:

The authors have made laudable efforts to address the criticisms I made in my evaluation of the original manuscript.

We thank the reviewer for their support!

Recommendations for the authors:

Reviewing Editor:

I would suggest two minor edits:

- The findings are correlative and descriptive, but the title implies functionality (A New Role for RNA G-quadruplexes in Aging and Alzheimer′s Disease). I would suggest toning down this title).

- While I understand the limitations in performing additional biochemical experiments to validate the immunofluorescence study, I think this is worth mentioning as a limitation in the text.

We have made these two changes as requested, altering the title to remove the word Role that may imply more meaning than intended, and adding a line to the discussion on the need for future additional biochemical experiments.

Reviewer #1 (Recommendations for the authors):

Thanks for addressing the concerns raised.

We thank the reviewer for their support!

Reviewer #2 (Recommendations for the authors):

Minor point:

Related to the "correlation is not causality" remark I made in my evaluation of the original manuscript: the authors' answer is reasonable. Still, I would suggest to modify the abstract: "we propose a model of neurodegeneration in which chronic rG4 formation drives proteostasis collapse" => "we propose a model of neurodegeneration in which chronic rG4 formation is linked to proteostasis collapse"

All other remarks I made have been answered properly.

We thank the reviewer for their support! We have made the change exactly as requested by the reviewer.

Version 2

Abstract

Abstract

Introduction

As the world population ages, new molecular targets in aging and Alzheimer’s Disease (AD) are needed to combat the expected influx of new AD cases. Until now, the role of RNA structure in aging and neurodegeneration has largely remained unexplored.

Methods

In this study, we examined human hippocampal postmortem tissue for the formation of RNA G-quadruplexes (rG4s) in aging and AD.

Results

We found that rG4 immunostaining strongly increased in the hippocampus with both age and with AD severity. We further found that neurons with accumulation of phospho-tau immunostaining contained rG4s, that rG4 structure can drive tau aggregation, and that rG4 staining density depended on APOE genotype in the human tissue examined.

Discussion

Combined with previous studies showing the dependence of rG4 structure on stress and the extreme power of rG4s at oligomerizing proteins, we propose a model of neurodegeneration in which chronic rG4 formation is linked to proteostasis collapse. These morphological findings suggest that further investigation of RNA structure in neurodegeneration is a critical avenue for future treatments and diagnoses.

Introduction

As the aged population increases, Alzheimer’s Disease (AD) and AD-related dementias (ADRDs) will become one of the world’s largest medical and economic crises due to the lack of early diagnostic methods and durable therapies1, 2. Age is the highest risk factor for AD, as disease prevalence increases exponentially as a function of age, and close to 40% of older adults over the age of 80 experience AD or ADRDs3, 4. The most important genetic risk factor for developing late onset AD (LOAD) is the Apolipoprotein E (APOE) genotype, at least in Caucasians, where carriers of one or two copies of the APOE4 allele have a substantial increase in the risk for developing LOAD when compared to carriers of the APOE2 or 3 alleles4, 5. Beyond APOE, recent meta-analyses combining genome-wide association studies (GWAS) have expanded the number of AD-risk loci6, but most of the disease-associated variants reside in non-protein coding regions of the genome, making it difficult to elucidate how they affect AD susceptibility and pathogenesis.

The exact mechanisms behind these two predominant AD risk factors remains a mystery but both are widely associated with increased abnormal protein aggregation, which is the most abundantly observed pathological hallmark of AD pathogenesis. AD and ADRDs all have an abnormal protein aggregation profile that includes: 1) extracellular amyloid senile plaque deposition occurring at least 5-10 years before mild cognitive impairments (MCI) are observed and, 2) aggregates of phosphorylated Tau (p-Tau) forming intracellular neurofibrillary tangles (NFTs) near MCI onset7, 8. Novel biomarker studies have shown that hyperphosphorylated Tau species exist in plasma prior to onset of dementia symptoms, with specific phosphorylation sites seeming to affect the affinity of binding of the Tau protein to microtubules 9. With advancing age, protein aggregation is accelerated by the accumulation of reactive oxygen species (ROS) as well as the onset of neuroinflammation in the brain10-12. Having APOE4 was shown to significantly increase accumulation of amyloid and NFTs in Caucasians, and also increases inflammation and reactive oxygen species in the human brain5, 13. The effects of these AD risk factors have been studied mostly at the protein level in the brain, but not yet at RNA level, where structure and function have a myriad of important roles in numerous biological processes.

RNA higher order structures of particular interest are RNA guanine-rich G-quadruplexes (rG4s)14. G-quadruplexes are secondary structures that can fold under physiological conditions by guanine-rich DNA and RNA molecules and occur in the genomes and transcriptomes of many species, including humans15. rG4s are four-stranded helices formed by guanine tetrads through Hoogsteen base-pairing, with a monovalent cation channel running down the middle of the tetrads (Fig. 1a)16. G4s are potentially involved in the regulation of DNA and RNA processes, such as replication, transcription, translation, RNA localization, and degradation17-19. Recently, rG4s were shown to form as a function of stress in eukaryotic cells20. Under non-stress conditions, rG4s were largely unfolded, but upon introduction of stress (including ROS stress) the rG4s preferentially folded and remained folded until the cessation of stress21. This observation raises the possibility that rG4s could preferentially fold under the increased and chronic stress conditions of aging and AD in the human brain, leading to abnormally high levels of rG4 formation in older adults or in adults with protein misfolding diseases22. RNA has been known to be included in AD-related aggregates of both amyloid and NFTs in patients with AD for decades23, and a recent sequencing study of nucleic acids embedded in AD aggregates showed an enrichment in potential rG4-forming sequences24.

a, (left) Representative G-quadruplex tetrad structure with Hoogsteen base pairing stabilized by a cation at the center. (right) G-quadruplex structure with stacked tetrads. b, Demographics table for individuals included in this study. c, Immunofluorescence showing BG4 staining (red) in brain sections of human hippocampus from four control individuals of increasing age ranging 30-91 years old (Case 15, 13, and 8, respectively) all with Braak stage 0. d, Correlation of percent area covered by BG4 fluorescence in the outer-molecular layer versus age of individual.

It has recently been shown that rG4s can greatly enhance protein oligomerization25, 26. Moreover, rG4s can bind to and greatly enhance the phase transition of tau27. Similarly, Zwierzchowski-Zarate et al explored the effects of different RNA sequences on tau aggregation in a cellular model28. In their study, several sequences caused an increase in tau aggregation, but the highest level of tau aggregation was achieved with a GGGC repeat sequence28. However, the structure of the RNAs, including possible rG4 formation, was not explored or speculated upon. We therefore performed CD spectroscopy to determine if the sequence associated with the highest levels of tau fibrillation might be forming rG4 secondary structures. The CD spectra show that this RNA forms parallel rG4 structures in vitro (Supplementary Data Fig. 1).

Based on these previous in vitro and cellular studies, an increased presence of rG4s in the brain could contribute to AD-associated aggregation. However, the presence of rG4s in the human brain or other human tissues has not been shown previously, nor how it relates to aging or AD pathology. This was the focus of the current study.

Methods

Human brain tissue

This work was approved by the University of Colorado Institutional Review Board (IRB) and the study was performed in accordance with the ethical standards of the 1964 Declaration of Helsinki and its later amendments. However, the use of postmortem brain tissue for research is exempt according to federal IRB regulations and does not require IRB review and approval. All efforts were made to consider diversity, equity and inclusion (DEI) in the selection of brain tissues for this study. The tissues included in this study were obtained from the Medical University of South Carolina (MUSC) brain bank (Dr. Eric Hamlett). Postmortem consent was obtained from the next of kin, and the brain was rapidly removed. Fig. 1b shows the demographics for the 21 cases used for this study. The tissue was sliced in 1 cm coronal slices and subjected to free-floating fixation in 4% paraformaldehyde in phosphate buffered saline (PBS) for 72 hours. The fixed tissue was then transferred to a 30% sucrose in phosphate buffer (PB), followed by transfer to a cryoprotection solution in PBS (30% Glycerol, 30% Ethylene Glycol and 40% PO4 Buffer) and stored in -20°C until dissection. Anatomical blocks from 19 brain regions were dissected from the 1 cm slabs, embedded in paraffin and cut at a thickness of 5µm on a Microm Microtome (Thermo Fisher Scientific Inc., Waltham, MA) and stained for routine neuropathological staging according to the NIA-AA Revised Criteria for Diagnosis and Staging of Alzheimer’s Disease 29.

Immunofluorescence

Paraffin-embedded sections were deparaffinized in xylene (RPI, CAT#111056) and rehydrated using wash steps at: 100%, 95%, 70%, 50% ethanol followed by distilled water. Heat-induced antigen retrieval was performed using 0.05% citraconic acid at pH 7.2 for 20 minutes in a steamer. Slides were then washed with a saline solution buffered with tris(hydroxymethyl) aminomethane (Tris, TBS) and blocked with 0.25% triton-X and 5% normal serum for 1 hour. After blocking, slides were rinsed again with TBS and incubated with 1X True Black Lipofuscin Autofluorescence Quencher (Biotium CAT#23007) diluted in 70% ethanol. Slides were washed 3 times with TBS then incubated with primary antibodies listed in Supplementary Data Table 1 overnight, followed by Goat anti-mouse AlexaFluor 555 (Thermo Scientific #A32727) or Donkey anti-rabbit Alexafluor 488 (Invitrogen #A-21206) at 1:500 dilution for 1 hour. Slides were then washed 3 times with TBS and cover-slipped using mounting media containing DAPI (Thermo Scientific, CAT#P36931). Staining controls included omission of primary or secondary antibodies (Supplementary Data Fig. 2). In Fig. 3b, we performed sequential double labeling, first with BG4, and then with rRNA antibody at 1:250 (Abcam #ab17119).

a, Immunofluorescence showing BG4 staining (red) in brain sections of human hippocampus with increasing Braak stage (1-6) associated with AD severity (Case 16, 11, 10, and 8, respectively). All scale bars are 300 µm. b, Correlation of percent area BG4 coverage in the outer molecular layer versus severity of Braak stage. c, Quantification of percent area BG4 coverage by population with E2/E3 and E3/E3 versus E3/E4 and E4/E4 APOE alleles.

Cellular types and localization of rG4s, all scale bars are 50 µm. a, Immunofluorescence showing BG4 staining patterns (red) in human hippocampal tissue of case 17, a 53-year-old individual with Braak stage 1.5, and case 4, a 79-year-old individual with Braak stage 5 AD pathology (right). Cell nuclei stained with DAPI (blue). b, Co-stain of rRNA (left, yellow green) and BG4 (right, red), in the hippocampal CA4 region, an 80-year old individual with early stage tauopathy, demonstrating significant but not complete overlap between BG4 and rRNA. c, BG4 staining (red) with cell type markers (green) of oligodendrocytes (SOX10, case 8), neurons (NEUN, case 4), microglia (IBA1, case 18) and astrocytes (GFAP, case 18). BG4 is most prevalent in oligodendrocytes, neurons, and astrocytes. BG4 does not strongly colocalize with microglia (IBA1).

In addition, we examined whether Braak stage correlated with aging itself (Supplementary Data Fig. 3). Although not significant at the 0.05 level, there was a weak correlation between Braak and Age. Braak staging is a paradigm used to describe density and regional spread of neurofibrillary tangles (NFTs) in the Alzheimer brain and includes levels 0-VI 7. CERAD (Consortium to Establish a Registry for Alzheimer’s Disease) is a staging paradigm for Alzheimer’s disease that takes into account amyloid plaque formation30.

Imaging and Statistical methods

Slides were imaged on a confocal microscope at 10X, 20X and 60X magnification at wavelengths 355, 488, and 555. Images used for quantification were taken at constant microscope and intensity settings. All image quantification was performed using Fiji ImageJ. Quantification was performed by a researcher who was blinded to the neuropathological diagnosis of individual cases and with no knowledge of Braak Stage or Age for the case being quantified. Supplemental Table 2 presents side by side comparisons of quantification from 2 researchers who were both blinded to group diagnosis and individual cases. Data used in the paper are highlighted. One data point with too much variation between quantifications was removed to increase rigor. Background was subtracted using a rolling ball radius. Images were then divided into 3 regions of interests (ROI); OML, DG and CA4. (Supplementary Data Fig. 4) Images were submitted to an auto-threshold and particle’s analyzed to get a percent area measurement of BG4 staining. These were graphed and linearly fit using Kaleidagraph using Pearson fit. Spearman and Pearson R and p-values are reported for 18 cases for age correlation and 18 cases for Braak correlation. An unpaired T-test with unequal variance was performed for the two groups E2/E3 and E3/E3 vs E3/E4 and E4/E4. When the singular African American case was removed from the APOE quantification, the t-probability remained highly significant at 0.0045 vs 0.002. All quantification was performed on 10X images obtained from an Olympus microscope.

Colocalization of rG4s with p-tau and neurodegeneration model.

a, Immunofluorescence of human hippocampus from 2 older AD individuals (case 8 and case 21)showing BG4 (red), DAPI (blue), and p-tau (green). b, Proposed model relating chronic rG4 formation and protein aggregation in aging and neurodegeneration.

Circular Dichroism

CD spectra were obtained using a Jasco J-1100 circular dichroism at 25°C. Sequence was resuspended in 10mM potassium phosphate pH 7.5 buffer and diluted to 12.5 µM (per strand) RNA. The CD measurement was taken from 300 nm to 210 nm at 1nm intervals using a 50 nm/min scanning speed. The sequence used was 5’-CGGGCGGCGGGGGGGCCCGGGCGGCGGGGGGGCCCGGGCG-3’ 28. Parallel structure was determined using previously described quantification methods for quadruplex topology31.

Results

Formation of rG4s in aging

To determine if rG4s form in the human brain, we used the well-characterized single-chain BG4 anti-quadruplex antibody (Sigma-Aldrich MABE917, antibody null controls in Supplementary Data Fig. 2) to stain human hippocampal postmortem tissue of different ages under conditions that strongly favor cytoplasmic rG4 identification using triton-X fixation in cells 32, 33. The tissues were obtained in collaboration with the brain bank at the Medical University of South Carolina (MUSC). The study included 21 cases, male and female, of different ages and with or without a clinical or neuropathological diagnosis of AD (see Fig. 1b). All cases had received a neuropathological examination including assessment of Braak and CERAD stages according to the updated NIA criteria except in those cases in which no pathology was noted 7, 34.

First, the rG4 staining profiles were compared between different ages, with an age range of 30-92 years of age (Fig. 1c). We discovered a significant age-related increase in rG4 immunostaining, with the densest staining observed in older individuals (Fig. 1c,d). The largest difference in staining between young and older cases was observed in the outer molecular layer (OML) and in the hilar region, CA4 (see Fig. 1c, legend arrows). The strongest overall staining for BG4 was found in the dentate gyrus granular cell layer (DG), but with less observable age-related increases. Staining density measurements confirmed these findings and demonstrated a highly significant increase in BG4 staining in the oldest cases studied within the OML (Fig. 1d) and the CA4 regions of cornu ammonis (Supplementary data Fig. 5a), confirming increased intracellular BG4 accumulation as a function of age. Plotting BG4 percent area versus age confirmed the observed findings and showed correlation (Spearman R= 0.49) with high significance (p= 0.038 in the OML region (Fig. 1d). Positive and significant Pearson but not Spearman correlations were observed between age and BG4 area in the CA4 region, Supplementary data Fig. 5a), and non-significant correlations were observed in the dentate granule cell layer. No significant correlation was found with BG4 stain and postmortem interval (PMI) (Spearman R = -0.18, p = 0.49 see Supplementary Data Fig. 6). In total, these data show that rG4s increased as a function of age in the human brain. We observed multiple outliers in both older and younger individuals due to Braak stage. For example, an 85-year-old individual with Braak stage 0, or 59-year-old individual with Braak stage 5.5 were outliers as a function of age. This led us to analyze whether rG4 formation increased with Braak stage, a measure of AD severity.

Formation of rG4s in AD

Next, we examined whether the formation of rG4s was associated with AD neuropathology. Braak staging is a neuropathological staging method that takes into account the density of neurofibrillary tangles (NFTs) both within each brain region and also from region to region30, with higher numerical values representing greater AD severity7. Like in the case of aging, increased BG4 staining was readily apparent in AD cases, especially in those cases associated with a higher Braak stage (Fig. 2a). In the OML, plotting BG4 percent area versus Braak stage demonstrated a strong correlation (Spearman R= 0.72) with highly significantly increased BG4 staining with higher Braak stages (p = 0.00086) (Fig. 2b). Positive significant correlations (R = 0.52, p = 0.028) between Braak stage severity and BG4 staining were also observed within the CA4 region (Supplementary data fig. 4). Interestingly, the OML region is highly involved in AD pathology, and is a hippocampal region where AD pathology is often observed,35 suggesting that the presence of rG4s in this particular brain region might have significance for amyloid as well as NFT formation during the AD disease process. In addition, a prominent loss of synapses in AD is found in the OML, suggesting a potential close connection with layer 2 of the entorhinal cortex36.

If rG4 formation is more prevalent in AD than in non-AD cases and directly involved in AD pathology, we would also predict that rG4 formation would be more prevalent in APOE genotypes that are predisposed for higher LOAD risk, i.e. patients with one or two copies of the APOE4 genotype. Indeed, there was a significant (t-probability 0.002) increase in BG4 staining levels in the OML in postmortem cases with one or more APOE4 alleles compared to those with only APOE2 or APOE3 alleles (Fig. 2c). This result, although correlative, is consistent with a previous study performed in vitro in which the RNA sequences found in aggregates in different APOE genotype cell lines were different24.

Patterns of rG4 localization

Images obtained at high magnification using a confocal microscope (Olympus) showed that a large amount of the BG4 immunostaining was localized in the cytoplasm near the nucleus in a punctate pattern, reminiscent of known rG4 formation in vitro and in cell lines in rRNA ribosomal extensions33. We therefore also stained with an anti-rRNA antibody to determine whether this was the case in human tissue as well. As expected, the stains overlapped (Fig. 3b, Fig. S7), showing that at least a large percentage of the observed BG4 signal arose from rG4s and not nuclear DNA G4s. This result also confirms that previously observed rG4 formation in ribosomes33 also occurs in human brain tissue.

To further explore the formation of rG4s in the human hippocampus, we examined the staining pattern in different cell types based on specific antibody markers. Co-staining with markers for the most common brain cell types, rG4s were clearly abundant in neurons, oligodendrocytes, and astrocytes (Fig. 3c). rG4s exhibited no apparent colocalization with microglia (Fig. 3c). This pattern is especially noticeable in the double labeling with SOX10 (Fig. 3c), which clearly shows the nuclear oligodendrocyte marker SOX1037 staining in proximity to the cytoplasmic BG4 immunostaining.

We co-stained BG4 with different p-tau epitope antibodies to determine whether pathological aggregation of tau proteins co-occurred with BG4 immunostaining (Fig. 4). Indeed, we found that BG4 immunostaining co-localized with hyper-phosphorylated tau immunostaining in neurons located in the CA1-4 regions of the hippocampus (Fig. 4). We used two different antibodies directed against the serine 396 site (S396) and the threonine 231 (T231) to investigate co-localization with different p-Tau epitopes. Quantifying the percentage of cells co-stained with BG4 and pTau revealed that 85% of the S396 p-tau positive cells also had positive BG4 staining, and 82% of the T231 p-tau positive cells also had positive BG4 staining.

Discussion

In this study, we demonstrated that rG4s form preferentially as a function of age and AD pathology in neurons and glial cells the human postmortem brain. While the granule cell layer in the dentate gyrus exhibited the most BG4 immunostaining overall, this was not an area most changed with aging and AD, as considerably greater changes were observed in the OML and the CA4 region of the hippocampus. rG4s were primarily found in the cytoplasm near the nucleus of neurons, astrocytes, and oligodendrocytes and were also found in neurons that contained regions of positive staining against two different epitopes of p-tau. The BG4 staining levels were also impacted by APOE status, with the presence of one or two APOE4 alleles associated with higher BG4 levels in the OML region. The OML is an interesting region of the hippocampus from the pathology standpoint, since this is the region of the hippocampus exhibiting the most prominent synaptic loss early in the disease process36, 38. We have recently demonstrated a significant loss of presynaptic components of the OML in AD38, strengthening the findings observed herein and suggesting a particular vulnerability of this region of the hippocampus to AD progressive pathology.

Combining these observations with those previously made on the reliance of rG4 structure on cellular stress and the roles of rG4s in protein oligomerization and tau aggregation in vitro, we can propose a model of how rG4s form and could perpetuate the effects of AD (Fig. 4c). As ROS and other protein stresses accumulate with age and according to APOE genotype, rG4s form. Although rG4s can rescue protein folding and could release proteins when unfolded at stress cessation39, under persistent stress proteins RNA would not be released from the G-quadruplexes, leading to increased protein oligomerization and aggregation26. This function has previously been suggested for RNA in biomolecular condensates more generally40. The acceleration of protein oligomerization and aggregation by the rG4s could then contribute to a vicious cycle of increased aggregation, leading to synaptic loss as described recently for the OML38. Recent in vitro work confirms an interaction between rG4 formation and p-Tau aggregation specifically41. In vitro experiments focused on the specific mechanisms for the rG4 influence on Tau phosphorylation and/or aggregation will be an important future focus following these preliminary observations to more clearly delineate roles of rG4s in neurodegeneration. Of note, the previous studies showing that the fixation conditions used here strongly favor rG4 identification instead of DNA G4 identification were performed in cells, and so it is possible that in tissue some DNA staining could also occur, however this is not suggested by the co-stain with rRNA (Fig. 3b).

Of note, the accumulation or rG4s in neurodegeneration could have roles in addition to those of protein aggregation. RNA transport dysregulation has also been a common factor identified in neurodegeneration42, and it could be possible that the rG4 build-up observed here could be in part a product of this process or could lead to it. It has also been proposed that the RNA transport dysregulation and protein aggregation processes could be linked 43. This remains an area of needed future investigation.

This work raises the possibility of using rG4s as potential drug targets in AD, or given the high correlation with AD severity, as early AD biomarkers. Future work can focus on ameliorating AD symptoms through rG4 binding molecules or detecting brain-derived rG4s as disease biomarkers for diagnosis. Many different rG4-binding molecules have already been created and shown to bind in human cells44, and provide a set of promising lead compounds for potential AD treatment.

Although here we have examined only AD, there are many other aging-related diseases featuring protein aggregation, including Parkinson’s Disease, ALS, Fragile X syndrome, and Huntington’s Disease 45. Notably, all of these diseases now have significant evidence showing rG4s as interactors with the aggregating proteins 46-58, and in the case of Fragile X, a recent study demonstrated that rG4s could contribute to its aggregation and neurotoxicity59. In addition, a recent manuscript by Raguseo et al., showed an increased prevalence of G4-structures in C9orf72 mutant human motor neurons in ALS/FTD when compared to healthy motor neurons, and direct involvement of the rG4 complex in the aggregation process60. These recent findings illustrate the need to explore rG4 complexes and their role in neurodegeneration further. The results here combined with these previous studies suggest that the formation of rG4s could be a common mechanism of neurodegeneration.